\textbf{\sc Towards Deliberative Control in Marine Robotics}

### 2012-05-04 Review-response.tex

\input{preamble}

\renewcommand{\baselinestretch}{0.973}

\begin{document}
\thispagestyle{empty}
\pagestyle{empty}
\setlength{\parindent}{0pt} 
\setlength{\parskip}{2ex}

\small{Kanna Rajan}\\
\small{Principal Researcher for Autonomy}\\
\small{Monterey Bay Aquarium Research Institute}\\
\small{Moss Landing, California}\\
\small{http://www.mbari.org/staff/kanna/}\\
\small{http://www.mbari.org/autonomy/index.htm}\\
\small{(831) 775-1833}\\

@@

Dear Mae,

First and foremost, we would like to thank the reviewer of our
manuscript above for the extensive and valuable comments provided. And
we thank you for your efforts to pull this all together.

We are heartened by the reviewers comments that the manuscript
``follows the intent and spirit'' of the book's theme. And we are
delighted in her/his belief that the manuscript should be accepted for
publication.

We have \emph{carefully} gone over all the comments provided. The
attached revised manuscript reflects those questions/comments from the
review and are highlighted in \kcomment{red} . In a number of
instances, we have taken the liberty to tighten the text for the
English and in one instance added some suplemental material which is
substantial, to explain how the \eu planner components work. We have
also added a couple of figures to better illustrate the concepts per
the reviewer's suggestions. The changes in sum are \underline{not}
substantial given the feedback. However the page count has, with the
changes, gone up a little. And as you would likely see, we've worked
hard to address all the comments while keeping the material
accessible.

@@

Thank you once again for all your efforts on our collective behalf for
this book. We look forward to working with you to bring the book out
for educators at the earliest.

\vspace{+2cm}
\begin{tabbing}
\hspace{+10cm}\=\small{Kanna Rajan}\\
\>\small{Corresponding Author}\\
\end{tabbing}

### 2012-05-04 arch.tex
updates

Systematic partitioned inference is a major contribution of
\rxe. However that is not the focus of this chapter with details
covered in \cite{py10, rajan12}. Coupled interaction between planning
and the world on the other hand, has a substantial impact on how
planning can be integrated in such frameworks. In this chapter we show
how using the \eu planner can be coupled to the \rx framework by first
linking concepts in temporal planning with the framework. Note that in
the following treatise, we call the formal semantic framework of \rx
as the 'agent'. Reactors constitute an entity within this agent and
are standalone blocks of inference. The scope, detail and number of
reactors are application specific and outside the scope of this
chapter.

@@

\begin{figure}[!htbp]
  \centering
  \includegraphics[scale=0.35]{figs/token-disptch.pdf}
  \caption{\small \kcomment{Dispatching a token occurs only when the request
    start time overlaps a window. \texttt{Goal$_1$} therefore gets
    dispatched from \texttt{reactor$_1$} to \texttt{reactor$_2$} but
    not \texttt{Goal$_2$} to avoid early commitment.}}
  \label{fig:reactor-lookahead}
\end{figure}

This window is used by \rx to identify when to dispatch a request to a
{\em reactor} as a new goal. This happens only if the request start
time -- represented as the interval of possible start times --
intersects this window. Requests further in the future are buffered by
the agent until this time window advances for an overlap. This
provides an elegant mechanism to request a change from the {\em
  external} state variable of a reactor while avoiding flooding the
owner reactor of this state variable and avoiding early commitment.
\kcomment{Fig. \ref{fig:reactor-lookahead} illustrates an example
  between two reactors where one token is dispatched when overlapping
  the dispatch window while another is not.}

The choice of both the latency and look-ahead of a reactor can be
delicate as one is dependent on the other. Needless to say, latency is
highly dependent on exogenous paramaters such as the host computers
computing power and the number of reactors needing to deliberate at
once, among others. As the design stands, these variables should be
evaluated carefully by the system designer balancing agent efficacy.

@@

Given the above example, \rx abstracts reactor interaction by
constraining them to state variables. Each reactor is therefore
agnostic where its {\em external} state information is managed as long
as this information is available and properly maintained for managing
{\em internal} state information. % This {\em internal} state in turn
% may be used by other reactors.
Such a formalism also constrains reactor interaction to state
information exchange which comes in two forms as illustrated in
Fig. \ref{fig:trex:flow}:

@@

\item {\em Observation on current state values}: This information
  propagates \emph{up} in the reactor dependency graph at the
  synchronization phase and occurs at the beginning of every
  tick. During this phase the agent provides updates to each reactor
  on their {\em external} state variables for this tick so they can
  compute their {\em internal} state variable values for this same
  tick. By doing so we propagate a consistent view of that state of
  the world throughout all the reactors as time
  advances. \kcomment{Observations can therefore be both sensor
    measurements as well as state information, passed up from other
    reactors. The strict rules of timeline ownership enable a clear
    policy for conflict resolution: observations dominate
    expectations.}

\item {\em Goal request on future state}: This information propagates
  \emph{down} in the reactor dependency graph. In order to satisfy own
  {\em internal} objectives, a reactor may have a plan that relies on
  future values of one of its {\em external} state. When such a plan
  is identified the agent ensures that this part of the plan is
  provided as a goal to the owner of the given {\em external} state
  variable(s).

@@

Both these functions have different scope and temporal constraints.
The \texttt{synchronize} call is considered as atomic and will focus
only on state inference for the current tick by identifying the
current value of {\em internal} state variables from the latest
updates on {\em external} state variables. This process propagates up
in the reactor dependency graph at every tick to ensure that all
reactors are aware of the current state of the world. \texttt{step}
relates to a single decision step of the planning process focused on
the future evolution of state variables; it can take multiple steps
spanning multiple ticks for a reactor to produce a plan that is both
complete and valid. How the reactor identifies that its plan is
complete and valid is its responsibility and depends on the way each
reactor resolves its plan at each {\em step}. The general idea is that
the reactor is able to find a valid strategy in order to satisfy as
many of its {\em internal} goals as possible. For \eue, this means
that the solver is unable to identify any further flaws to be resolved
in the plan.

When such plan is identified, its {\em external} timeline(s) are then
dispatched to the corresponding reactors which own these times and
which in turn can start their own deliberation \texttt{step}s to
satisfy these new objectives. Consequently execution occurs as these
goals and plans propagate down in the reactor hierarchy until they
eventually become goals for the lowest level reactor which -- instead
of deliberating -- may just transform this goal into a command for
execution in hardware.

\label{sec:delib-exec-as}
\begin{figure}[!htbp]
  \centering
  \vskip-1pc
  \subfloat[\small \kcomment{A view of synchronization and deliberation within a
  tick cycle.}]{\label{fig:sync-cycle}\includegraphics[width=0.35\columnwidth]{figs/sync-tick-cycle}} 
  \qquad
  \subfloat[\small A \rx reactor's execution cycle: {\em Deliberation} is
    interrupted by {\em synchronization} at the beginning of every
    {\em tick} allowing integration of state information.]{\label{fig:tick-exec}\includegraphics[width=0.45\columnwidth]{figs/tick-cycle}}
  \caption{\small \kcomment{Detailed and abstracted views of
      synchronization and deliberation within one tick cycle.}}
  \label{fig:sync-plan-cycle}
\end{figure}

@@

Within \rxe, an instance of an \eu reactor comes with the sole purpose
to deliberate. A typical design of such a reactor takes its cues from
\texttt{IDEA} \cite{mus02, mus06} where both planning and execution
are based on a single unified model within a rich representation that
provides support for temporal inference. In \rxe, the execution cycle
consists of two deliberation processes which are implemented as two
separate \eu solvers modifying the \emph{same} plan database.

@@

\item The \emph{synchronization solver} is a specialized \eu solver
  that integrates new information in the plan. This signifies the
  evolution of \emph{external} timelines and ensures that the reactor
  propagates these to identify current reactor state as well as to
  inform other reactors of any state change on its \emph{internal}
  timelines. This solver is summoned at the beginning of every single
  tick.

\item The \emph{deliberation solver} manages the deliberation process
  of the reactor. It does so either to produce a new plan, to alter
  its current plan as new goals are provided or when the
  synchronization solver identifies a conflict between current state
  and expectations from a previously generated plan. This process can
  span multiple ticks and can therefore be interrupted by the
  synchronization solver.

@@

While the two processes are separate, they share the same plan
internal to a reactor; consequently at every synchronization cycle the
planning process is informed of the new world state and its impact on
the plan. Conversely, when synchronization occurs after deliberation
stops, its starting point is the last partial plan resulting from
deliberation. This gives an initial frame to the problem that helps to
target the synchronization problem by ensuring that the new {\em
  external} state updates are compatible with this plan. If so,
synchronization propagates these events in the plan and can use it as
an aid in resolving its {\em current state flaw}. It does so by using
the plan information on the expected {\em internal} state evolution
near the execution frontier.

@@

This assumption, while helpful to reduce the scope and complexity of
the planning problem, becomes problematic for planning within a highly
dynamic environment. In a typical robot this implies that the only
point where the planner can be informed about the current world state
is prior to the commencement of a planning cycle, by providing the
planner an initial state and current objectives. During the planning,
it is assumed that the agent can maintain the world as perceived by
the planner. \cite{lemai04, lemai-chenevier2004} attempts to reduce
the impact of planning by allowing local plan repair in the existing
plan. In the eventuality that such repair is not feasible they require
that their robot comes to a stop until (re-)planning is complete.

This assumption is problematic for situated agents in general and as
Fig. \ref{fig:tick-exec} shows, for \rxe. Synchronization that tracks
state evolution can occur several times during reactor
deliberation. By making synchronization and deliberation share the
same plan we provide a solution that relaxes this static world
assumption. \texttt{IDEA} uses the same principle with all solvers in
an agent sharing the same plan for communicating environmental change
\cite{Dias:2003ua, mus06}. Specifically their {\em Reactive planner}
-- similar to our {\em synchronization} solver -- was often the solver
targeting the change at the execution frontier. 
% Although, none of the papers we found were detailing how this was
% handled in the architecture.
\cite{Dias:2003ua} states:

@@

This suggests a deliberate non overlap between \kcomment{the} {\em
  reactive planner} and {\em deliberative planners} horizons. In \rxe,
and more specifically in our \eu reactor the planning horizon of {\em
  synchronization} -- limited to the current tick -- is a subset of
the planning horizon of the deliberation steps -- starting at the
current tick and with an extent based on latency $\lambda$ and
look-ahead $\pi$ of the reactor. In this manner we avoid creating a
conflict between the deliberation and synchronization by controlling
the scheduling of these two process\kcomment{es} by the \rx agent.

In the following sections we show how synchronization and deliberation
are implemented independently to show how their interaction through
the plan helps relax this assumption. It does so by integrating
planning in the reactor execution cycle even while allowing for more
informed deliberation.

@@

We first show how a reactor can track and identify its state in the
context where it does not have any compelling need to deliberate. In
doing so we assume that we have a reactor that has no future
goal. Under such a condition, \rx requires each reactor to produce at
every tick its {\em internal} state. For those reactors not owning
{\em internal} timelines, \rx makes sure that {\em external}
observations are taken into account for every tick as shown in
\cite{py10}. Such a requirement is enforced to ensure that all
reactors share a consistent view of the world with advancing time.

{\em External} state updates are managed by introducing these
observations in the plan structure as facts and ensuring that they
remain compatible with the current plan maintained by the reactor. The
challenging part of synchronization is to decide the {\em internal}
state values for the current tick. In the Shopping example for example,
we have the following code snippet which helps identify when we can
\texttt{Buy} a product:

\begin{verbatim}
Agent::Buy {
  eq(duration, 10);
  // [...]
  contained_by(condition object.location.At currLocation);
  // [...]
}
\end{verbatim}

@@

In \texttt{T-REX}'s \eu reactor this decision-making requirement is a
new type of flaw which requires identification of all values of {\em
  internal} state variables for the current tick. This flaw can be
defined as:

@@

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=0.5\columnwidth]{figs/synch-relation}
  \caption{\small Illustration of synchronization flaws in a
    reactor. The reactor receives new observations when they are
    produced by the owner(s) of its internal timelines. The open
    ``holes'' after the last token of each timeline represents the
    domain of all possible values for the end of this token. At every
    tick $\tau$ the reactor needs to integrate the {\em External}
    state information it received and -- based on its model
    $\mathcal{M}$ -- resolve its {\em Internal} state that will then
    be provided by the agent to other reactors using these state
    variables.}
  \label{fig:synch:flaw}
\end{figure}

@@

Intuitively the first choice enforces an \emph{inertial value
  assumption} where the reactor will tend to maintain its current
state as long as the model and external observations allow for it. The
second choice favors the advance of the reactor's current plan if any;
this directs the {\em synchronization} decision toward assuming that
its current plan is properly executed. Finally if none of these two
choices provides a valid solution -- \ie lead to an inconsistant state
-- the reactor tries all the possible state values remaining for the
timeline until it finds one that brings plan consistency.

% allows it to identify its state as required by \rx architecture.

% The flaw forces the reactor to identify its internal state for the
% current tick. By using this flaw one can describe the synchronization
% process as:

@@

\item Integrate the external state as provided by the owner of each
  external timeline into the plan database. 
  \begin{itemize}
  \item New observations are integrated by creating the
    corresponding token as a fact in the plan database
  \item If no observations have been received at this tick the last
    observation token restrict its \texttt{end} domaine to the
    interval $[\tau+1, +\infty)$
  \end{itemize}

\item Propagate this information in the plan database using model
  $\mathcal{M}$ of this reactor.

\item Resolve the current state value of each internal timelines based
  on Proposition \ref{prop:csf:resolve}.

@@

Steps $2$ \& $3$ above are handled by the \eu solver from Algorithm
\ref{alg:europa:solve} which ensures that the {\em current state
  flaws} are selected last so as to follow the sequence given above.

As {\em synchronization} focuses on the execution frontier it also
filters out all flaws that are outside of the current tick's scope.
If the flaw correlates to a token that is {\em necessarily} ending
before $\tau$ or starting after $\tau+1$ then the solver ignores
it. Such a strategy helps to reduce the scope of synchronization
exclusively to the current state of the reactor and in turn reduces
the search space for the {\em synchronization} solver.  To do so we
need to assume that the current state value of an {\em internal}
timeline does not depend on the future or more accurately that any
choice made during this {\em synchronization} will not lead to a
future inconsistency.  Such an assumption impacts the set of possible
domains the reactor can support while remaining complete. Take for
example this rule from our \texttt{Shopping} model:

@@

In the real world the fact that the robot is \texttt{Going} to a
certain location does not necessarily imply that it will succeed in
ending up there. Due to an exogenous event, for example, the road to
the \texttt{SuperMarket} could be closed, which could result in an
alternate observation that the robot is \texttt{At(7th st. \& Main)}
contradicting the existing plan. Further, it is important to
understand the temporal direction of the \emph{causal link} that is
triggering the inconsistency. At the time the robot observes the
contradiction of the \texttt{At} predicate, the rule that generated
this conflict is connected to past observations \texttt{Going} and
\texttt{Go}. These cannot be retracted easily without a potential
impact to the plan implying in terms of a cost in back-propagation.

@@

One could therefore alter this part of the model to give the reactor
more flexibility as follows:

@@

We do so by relaxing the constraint in line $5$ from a strict
\texttt{equals} to a \texttt{contains} and to add the new constraint
at line $11$ which ensures that the last \texttt{Going} ends up in the
location the agent actually wants the robot to be at. This first level
of relaxation will therefore allow the agent to use multiple
\texttt{Going} tokens should the first one fail to reach the
objective.

@@

One critical aspect to take into account is that from a \rx point of
view, synchronization is the only critical task of a reactor. Such a
process is used in order for the agent to identify its current state
and maintain it consistently throughout the agent's lifetime. For a
reactor not being able to find a consistent solution during
synchronization -- implying that after evaluating all possible
solutions provided by a model it was unable to find a consistent plan
-- not only indicates that a reactor is unable to explain the state of
the world, but more importantly it may jeopardize other reactors that
rely on its {\em internal} state. For this reason a failure to
synchronize immediately results in the \rx agent killing a reactor and
notifying any reactor that depends on its {\em internal} state
variables; it does so by posting the special observation
\textsf{Failed}. This special state is a special predicate which is
acceptable for all the {\em internal} and {\em external} timelines of
\kcomment{an} \eu reactor.  By so doing we allow for graceful
degradation where other reactors can readapt their plans on receipt of
this observation.

Finally, at the engine level careful choices were made in both the
focus of the search -- limited only to tokens that can potentially
overlap the current tick -- and heuristics where the {\em current
  state flaw} is considered last during the search process. The later
limits the amount of backtracking for a solution -- or at least how
deep one needs to backtrack in a decision tree.  Similarly the
sequence in which we evaluate the possible resolution for such a flaw
-- presented in Proposition \ref{prop:csf:resolve} -- was selected
according to some simple assumptions, that any external observation
that is compatible with an existing token in the current plan reflects
the execution of the plan itself. In other words this means that our
solver always prefers to use tokens \emph{already existing} in the
plan as a way to identify current internal state. By doing so
synchronization exploits the structure of the plan produced during
previous deliberation steps.

@@

Deliberation inside a reactor is based on \eu capabilities described
in Section \ref{sec:europa:arch}. However this deliberation comes with
some augmentation, namely the reactor's deliberation latency
\kcomment{($\lambda$)} and its planning look-ahead \kcomment{($\pi$)}
both expressed in ticks. \kcomment{In addition,} the agent needs to
interleave deliberation with not just synchronization but also with
the deliberation of other reactors\footnote{The current implementation
  of \rx is running on a single process and it is the responsibility
  of the agent itself to emulate reactor multi-threading for
  deliberation and synchronization.}.  For this reason while the
reactor relies on the \eu solver, it slices the execution of Algorithm
\ref{alg:europa:solve} into atomic steps. This interruption occurs
around the recursive call line \ref{li:recurse} while ensuring that
the current partial plan is not inconsistant at any single step. This
is managed by allowing the call to backtrack until a
\textsf{DecisionPoInt} that is not exhausted (\ie $decision.hasNext()$
is \texttt{true} \kcomment{and non empty}) is found in the call
stack\footnote{\kcomment{The call stack refers to the sequence of
    recursive function calls.}}.

A critical aspect of this deliberation step is to ensure that at the
exit of this call the $plan$ is not inconsistent. % Indeed, it is always
% possible that the next call made by the agent is for synchronization
% which could fail if the $plan$ is inconsistent before any attempt is
% made for synchronization.
If the partial plan is found to be inconsistent during a deliberation
step we either backtrack in the decision stack until a node of the
\kcomment{search} tree is found with an alternate solution or the plan
is relaxed since there is no other alternative solution. We do this by
removing all goals while keeping only recent observations.

{\em Synchronization} and the dispatching of new {\em goals} from \rx
can occur between two steps and alter the plan. These alterations
exogenous to the deliberation process dynamically impact the set of
flaws that need to be resolved. New flaws are created when a new goal
is received from \rx and conversely the recall of an objective could
remove the corresponding flaws\footnote{However if the flaws related
  to this goal were resolved, the removal of this goal may likely
  create new flaws.}. The impact of {\em synchronization} can be more
subtle as its alteration of the plan at the execution frontier is more
drastic. We can still give a non exhaustive list of possible
alterations of the flaws directly related to {\em synchronization}:

@@

\item The {\em synchronization} solver may have created new flaws that
  were outside of its scope but are now part of the scope of the {\em
    deliberation} solver.

>>>>

Systematic partitioned inference is a major contribution of \rx
\kcomment{which is covered in more detail} in \cite{py10,
  rajan12}. Coupled interaction between planning and the world on the
other hand, has a substantial impact on how planning can be integrated
in such frameworks. In this chapter we show how using the \eu planner
can be coupled to the \rx framework by first linking concepts in
temporal planning with the framework. Note that in the following
treatise, we call the formal semantic framework of \rx as the
'agent'. Reactors constitute an entity within this agent and are
standalone blocks of inference. The scope, detail and number of
reactors are application specific and outside the scope of this
chapter.

@@

This window is used by \rx to identify when to dispatch a request to a
{\em reactor} as a new goal. This happens only if the request start
time -- represented as the interval of possible start times --
intersects this window. Requests further in the future are buffered by
the agent until this time window advances for an overlap. This
provides an elegant mechanism to request a change from the {\em
  external} state variable of a reactor while avoiding flooding the
owner reactor of this state variable \kcomment{and avoiding early
  commitment.} The choice of both the latency and look-ahead of a
reactor can be delicate as one is dependent on the other. Latency is
dependent on exogenous paramaters such as the \kcomment{computer power
  of the host} and the number of reactors needing to deliberate at
once, among others. As the design stands, these variables should be
evaluated carefully by the system designer balancing agent
efficacy. Fig. \ref{fig:reactor-lookahead} illustrates an example
between two reactors where one token is dispatched when overlapping
the dispatch window while another is not.}

@@

Given the above example, \rx abstracts reactor interaction by
constraining them to state variables. Each reactor is therefore
agnostic where its {\em external} state information is managed as long
as this information is available and maintained for managing {\em
  internal} state information.  Such a formalism also constrains
reactor interaction to state information exchange which comes in two
forms as illustrated in Fig. \ref{fig:trex:flow}:

@@

\item {\em Observation on current state values}: This propagates
  \emph{up} in the reactor dependency graph \kcomment{during} the
  synchronization phase and occurs at the beginning of every
  tick. During this phase the agent provides updates to each reactor
  on their {\em external} state variable(s) to \kcomment{enable
    computation of} their {\em internal} state variable values for
  this same tick. By doing so we propagate a consistent view of world
  state throughout the reactors as time advances. \kcomment{Further,
    observations can be both sensor measurements as well as state
    information, passed up from other reactors. The strict rules of
    timeline ownership enable a clear policy for conflict resolution:
    observations dominate expectations.}

\item {\em Goal request on future state}: This information propagates
  \emph{down} in the reactor dependency graph. In order to satisfy its
  own {\em internal} objectives, a reactor may have a plan that relies
  on future values of one \kcomment{(or more)} of its {\em external}
  state. When such a plan is identified the agent ensures that this
  part of the plan is provided as a goal to the owner of the given
  {\em external} state variable(s).

@@

Both these functions have different scope and temporal constraints.
The \texttt{synchronize} call is considered as atomic and focuses on
state inference for the current tick by identifying the current value
of {\em internal} state variables from the latest updates on {\em
  external} state variables. This process propagates up in the reactor
dependency graph at every tick to ensure that all reactors are aware
of current world state. \texttt{step} relates to a single decision
step of the planning process focused on the future evolution of state
variables; it can take multiple steps spanning multiple ticks for a
reactor to produce a plan that is both complete and valid. How the
reactor identifies that its plan is complete and valid is its
responsibility and depends on the way each reactor resolves its plan
at each {\em step}. The general idea is that the reactor is able to
find a valid strategy in order to satisfy as many of its {\em
  internal} goals as possible. For \eue, this means that the solver is
unable to identify any further flaws to be resolved in the
plan. \kcomment{Fig. \ref{fig:sync-cycle} illustrates the distinction
  between a tick, \texttt{step} and synchronization within the tick.}

When such a plan is identified, \kcomment{the reactors} {\em external}
timeline(s) are then dispatched to their \kcomment{respective reactor
  owners,} which in turn can start their own deliberation
\texttt{step}s to satisfy these new objectives. Consequently execution
occurs as these goals and plans propagate down in the reactor
hierarchy until they eventually become goals for the lowest level
reactor which -- instead of deliberating -- may just transform this
goal into a command for execution in hardware.

% \label{sec:delib-exec-as}
\begin{figure}[!htbp]
  \centering
  \vskip-1pc
  \subfloat[\small \kcomment{A view of synchronization and deliberation within a
  tick cycle.}]{\label{fig:sync-cycle}\includegraphics[width=0.35\columnwidth]{figs/sync-tick-cycle}} 
  \qquad
  \subfloat[\small A \rx reactor's execution cycle: {\em Deliberation} is
    interrupted by {\em synchronization} at the beginning of every
    {\em tick} allowing integration of state information.]{\label{fig:tick-exec}\includegraphics[width=0.45\columnwidth]{figs/tick-cycle}}
    \caption{\small \kcomment{Detailed and abstracted views of
        synchronization and deliberation within one and multiple tick
        cycles.}}
  \label{fig:sync-plan-cycle}
\end{figure}

@@

Within \rxe, \kcomment{the sole purpose of an instance of an \eu
  reactor is to deliberate}. A typical reactor design takes its cues
from \texttt{IDEA} \cite{mus02, mus06} where both planning and
execution are based on a single unified model within a rich
representation that provides support for temporal inference. In \rxe,
the execution cycle consists of two deliberation processes which are
implemented as two separate \eu solvers modifying the \emph{same} plan
database.

@@

\item The \emph{synchronization solver} is a specialized \eu solver
  that integrates new information in the plan. This signifies the
  evolution of \emph{external} timelines and ensures that the reactor
  propagates \kcomment{changes} to identify current state as well as
  to inform other reactors of any state change on its \emph{internal}
  timelines. This solver is \kcomment{invoked} at the beginning of
  every single tick.

\item The \emph{deliberation solver} is used to either produce a new
  plan, to alter its current plan as new goals are provided or when
  the synchronization solver identifies a conflict between current
  state and expectations from a previously generated plan. This
  process can span multiple ticks and can therefore be interrupted by
  the synchronization solver.

@@

While the two processes are separate, they share the same plan
internal to a reactor; consequently at every synchronization cycle the
planning process is informed of the new world state and its impact on
the plan. Conversely, when synchronization occurs after deliberation
stops, its starting point is the last partial plan resulting from
deliberation \kcomment{(see Fig. \ref{fig:sync-plan-cycle})}. This
helps to target the synchronization problem by ensuring that the new
{\em external} state updates are compatible with this plan. If so,
synchronization propagates these events in the plan and can use it as
an aid in resolving its {\em current state flaw}. It does so by using
the plan information on the expected {\em internal} state evolution
near the execution frontier.

@@

This assumption, while helpful to reduce the scope and complexity of
the planning problem, becomes problematic for planning within a
dynamic environment. In a typical robot this implies that the only
point where the planner can be informed about the current world state
is prior to the commencement of a planning cycle, by providing the
planner an initial state and current objectives. During the planning,
it is assumed that the agent can \emph{maintain the world as perceived
  by the planner}. \cite{lemai04, lemai-chenevier2004} attempts to
reduce the impact of planning by allowing local plan repair in the
existing plan. In the eventuality that such repair is not feasible
they require that their robot comes to a stop until (re-)planning is
complete.

This assumption is problematic for situated agents in general and as
Fig. \ref{fig:tick-exec} shows, for \rxe. Synchronization that tracks
state evolution can occur several times during reactor
deliberation. By making synchronization and deliberation share the
same plan we provide a solution that relaxes this static world
assumption. \texttt{IDEA} uses the same principle with all solvers in
an agent sharing the same plan for communicating environmental change
\cite{Dias:2003ua, mus06}. Specifically their {\em Reactive planner}
-- similar to our {\em synchronization} solver -- is the solver
targeting change at the execution frontier.  \cite{Dias:2003ua}
states:

@@

This suggests a deliberate non overlap between \kcomment{the} {\em
  reactive planner} and {\em deliberative planner}'s horizons. In
\rxe, and more specifically in our \kcomment{deliberative} reactor,
the planning horizon of {\em synchronization} -- limited to the
current tick -- is a subset of the planning horizon of the
deliberation steps -- starting at the current tick and with an extent
based on latency $\lambda$ and look-ahead $\pi$ of the reactor. In
this manner we avoid creating a conflict between the deliberation and
synchronization by controlling the scheduling of these two
process\kcomment{es} by the \rx agent.

The following sections detail how synchronization and deliberation are
implemented independently to show how their interaction through the
plan helps relax this assumption. It does so by integrating planning
in the reactor execution cycle even while allowing for more informed
deliberation.

@@

We first show how a reactor can track and identify its state in the
context where it does not have any compelling need to deliberate. In
doing so we assume that we have a reactor that has no future
goal. Under such a condition, \rx requires each reactor to produce at
every tick its {\em internal} state. For those reactors not owning
{\em internal} timelines, \rx makes sure that {\em external}
observations are taken into account for every tick as in
\cite{py10}. Such a requirement is enforced to ensure that all
reactors share a consistent view of the world with advancing time.

{\em External} state updates are managed by introducing these
observations in the plan structure as facts and ensuring that they
remain compatible with the current plan maintained by the reactor. The
challenging part of synchronization is to decide the {\em internal}
state values for the current tick. In the Shopping example from
\kcomment{Section \ref{sec:europa:modeling}} for example, we have the
following code snippet which helps identify when we can \texttt{Buy} a
product:

\begin{verbatim}
Agent::Buy {
  eq(10, duration);
  // [...]
  contained_by(condition object.location.At currLocation);
  // [...]
}
\end{verbatim}

@@

In the \kcomment{deliberative} reactor this decision-making
requirement is a new type of flaw which requires identification of all
values of {\em internal} state variables for the current tick. This
flaw can be defined as:

@@

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=0.5\columnwidth]{figs/synch-relation}
  \caption{\small Illustration of synchronization flaws in a
    reactor. The reactor receives new observations when they are
    produced by the owner(s) of its internal timelines. The open
    ``holes'' after the last token of each timeline represents the
    domain of all possible values for the end of this token. At every
    tick $\tau$ the reactor needs to integrate the {\em External}
    state information it received and -- based on its model
    $\mathcal{M}$ -- resolve its {\em Internal} state that will then
    be provided \kcomment{in turn} by the agent to other reactors.}
  \label{fig:synch:flaw}
\end{figure}

@@

Intuitively the first choice enforces an \emph{inertial value
  assumption} where the reactor will tend to maintain its current
state as long as the model and external observations allow for it. The
second choice favors the advance of the reactor's current plan if any;
this directs the {\em synchronization} decision toward assuming that
its current plan is properly executed. Finally if none of these two
choices provides a valid solution -- \ie \kcomment{it leads} to an
inconsistant state -- the reactor tries all the possible state values
remaining for the timeline until it finds one that \kcomment{results
  in} plan consistency.

@@

\item Integrate the external state as provided by the owner of each
  external timeline into the plan database.

  \item New observations are integrated by creating the
    corresponding token as a fact in the plan database

  \item If no observations have been received at this tick the last
    observation token restricts its \texttt{end} domain to the
    interval $[\tau+1, +\infty)$

\item \label{itm2} Propagate this information in the plan database
  using reactor model $\mathcal{M}$.

\item \label{itm3} Resolve the current state value of each internal
  timeline based on Proposition \ref{prop:csf:resolve}.

@@

Steps \ref{itm2} \& \ref{itm3} above are handled by the \eu solver
from Algorithm \ref{alg:europa:solve} which ensures that the {\em
  current state flaws} are selected last so as to follow the sequence
given above.

As {\em synchronization} focuses on the execution frontier it also
filters out all flaws that are outside of the current tick's scope.
If the flaw correlates to a token that is {\em necessarily} ending
before $\tau$ or starting after $\tau+1$ then the solver ignores
it. Such a strategy helps to reduce the scope of synchronization
exclusively to the current state of the reactor and in turn reduces
the search space for the {\em synchronization} solver.  To do so we
need to assume that the current state value of an {\em internal}
timeline does not depend on the future or more accurately that any
choice made during this {\em synchronization} will not lead to a
future inconsistency.  Such an assumption impacts the set of possible
domains the reactor can support while remaining
\kcomment{algorithmically} complete. Take for example this rule from
our \texttt{Shopping} model:

@@

In the real world the fact that the robot is \texttt{Going} to a
certain location does not necessarily imply that it will succeed in
ending up there. Due to an exogenous event, for example, the road to
the \texttt{SuperMarket} could be closed, which could result in an
alternate observation that the robot is \texttt{At(7th st. \& Main)}
contradicting the existing plan. Further, it is important to
understand the temporal direction of the \emph{causal link} that is
triggering the inconsistency. At the time the robot observes the
contradiction of the \texttt{At} predicate, the rule that generated
this conflict is connected to past observations \texttt{Going} and
\texttt{Go}. These cannot be retracted easily without a potential
impact on computational cost.

@@

One could therefore alter this part of the model to give the reactor
more flexibility by relaxing the constraint in line $5$ from a strict
\texttt{equals} to a \texttt{contains} and to add the new constraint
at line $11$ to ensure that the last \texttt{Going} ends up in the
location the agent actually wants the robot to be at. This first level
of relaxation will therefore allow the agent to use multiple
\texttt{Going} tokens should the first token fail to reach the
objective, as shown below:

@@

@@

Synchronization is the only critical task of a reactor. Such a process
is used in order for the agent to identify its current state and
maintain it consistently throughout the agent's lifetime. For a
reactor not being able to find a consistent solution during
synchronization -- implying that after evaluating all possible
solutions provided by a model \kcomment{being} unable to find a
consistent plan -- not only indicates that a reactor is unable to
explain the state of the world, but more importantly it may jeopardize
other reactors that rely on its {\em internal} state. For this reason
a failure to synchronize immediately, results in the \rx agent killing
a reactor and notifying any reactor that depends on its {\em internal}
state variables; it does so by posting the special observation
\textsf{Failed}. This special state is acceptable for all the {\em
  internal} and {\em external} timelines of \kcomment{an} \eu reactor.
By so doing we allow for graceful degradation where other reactors can
readapt their plans on \kcomment{receiving} this observation.

Finally, at the engine level careful choices \kcomment{are} made in
both the focus of the search -- limited only to tokens that can
potentially overlap the current tick -- and heuristics where the {\em
  current state flaw} is considered last during the search
process. The later limits the amount of backtracking for a solution --
or at least how deep one needs to backtrack in a decision tree.
Similarly the sequence in which we evaluate the possible resolution
for such a flaw -- \kcomment{described} in Proposition
\ref{prop:csf:resolve} -- \kcomment{is} selected according to
\kcomment{the assumption}, that any external observation that is
compatible with an existing token in the current plan reflects the
execution of the plan itself. In other words this means that our
solver always prefers to use tokens \emph{already existing} in the
plan as a way to identify current internal state. By doing so
synchronization exploits the structure of the plan produced during
previous deliberation steps.

@@

Deliberation inside a reactor is based on \eu capabilities described
in Section \ref{sec:europa:arch}. However this deliberation comes with
some augmentation, namely the reactor's deliberation latency
\kcomment{($\lambda$)} and its planning look-ahead \kcomment{($\pi$)}
both expressed in ticks. \kcomment{In addition,} the agent needs to
interleave deliberation with not just synchronization but also with
the deliberation of other reactors\footnote{The current implementation
  of \rx is running on a single process and it is the responsibility
  of the agent itself to emulate reactor multi-threading for
  deliberation and synchronization.}.  For this reason while the
reactor relies on the \eu solver, it slices the execution of Algorithm
\ref{alg:europa:solve} into atomic steps. This interruption occurs
around the recursive call, line \ref{li:recurse}, while ensuring that
the current partial plan is not inconsistant at any single step. This
is managed by allowing the call to backtrack until a
\textsf{DecisionPoInt} that is not exhausted (\ie $decision.hasNext()$
is \texttt{true} \kcomment{and non empty}) is found in the call
stack\footnote{\kcomment{The call stack refers to the sequence of
    recursive function calls.}}.

A critical aspect of this deliberation step is to ensure that at the
exit of this call the $plan$ is consistent.  If the partial plan is
found to be inconsistent during a deliberation step we either
backtrack in the decision stack until a node of the \kcomment{search}
tree is found with an alternate solution or the plan is relaxed since
there is no other alternative solution. We do this by removing all
goals while keeping only recent observations.

{\em Synchronization} and the dispatching of new {\em goals} from \rx
can occur between two steps and alter the plan. These alterations
exogenous to the deliberation process dynamically impact the set of
flaws that need to be resolved. New flaws are created when a new goal
is received from \rx and conversely the recall of an objective could
remove the corresponding flaws\footnote{However if the flaws related
  to this goal were resolved, the removal of this goal may likely
  create new flaws.}. The impact of {\em synchronization} can be more
subtle as its alteration of the plan at the execution frontier is more
drastic. \kcomment{A possible list of alterations of the plan directly
  related to {\em synchronization} could be}:

@@

\item The {\em synchronization} solver may have created new flaws that
  were outside of its scope but are now part of the {\em deliberation}
  solver's scope.

\item The last observation has proved that the current plan is invalid
  requiring the {\em synchronization} solver to relax the existing
  plan to identify its current {\em external} state. (This case is
  developed further \kcomment{in Section \ref{sec:arch:intertwine}}).
